---
title: "Giotto proximity enrichment"
output: html_document
date: "2024-07-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

# Load libraries

```{r}
library(Giotto)
library(Seurat)
library(reshape2)
library(ggplot2)
library(dplyr)
library(tidyverse)
library(viridis)
```

# Read data
```{r}
source('data_cleanup.R')
sp_data = spatial_clean()
a1 = sp_data[[1]]
a2 = sp_data[[2]]
c2 = sp_data[[3]]
d2 = sp_data[[4]]
```

# Run Giotto



```{r}
# Set python path for Giotto
instr <- createGiottoInstructions(python_path = "C:/Users/yunseolp/Documents/.virtualenvs/r-reticulate/Scripts/python.exe")
```

```{r}
# Set idents to annotations
Idents(a1) <- a1$annotationSave
Idents(a2) <- a2$annotationSave
Idents(c2) <- c2$annotationSave
Idents(d2) <- d2$annotationSave
```

## Create Giotto objects

```{r}
# A1-1
a1_giotto_obj <- createGiottoObject(GetAssayData(a1, layer="counts"),
                          norm_expr = GetAssayData(a1),
                          spatial_locs = a1@reductions$spatial@cell.embeddings,
                          cell_metadata = a1@meta.data,
                          instructions = instr)

#A1-2
a2_giotto_obj <- createGiottoObject(GetAssayData(a2, layer="counts"),
                          norm_expr = GetAssayData(a2),
                          spatial_locs = a2@reductions$spatial@cell.embeddings,
                          cell_metadata = a2@meta.data, instructions = instr)

#C2
c2_giotto_obj <- createGiottoObject(GetAssayData(c2, layer="counts"),
                          norm_expr = GetAssayData(c2),
                          spatial_locs = c2@reductions$spatial@cell.embeddings,
                          cell_metadata = c2@meta.data, instructions = instr)

#D2-1
d2_giotto_obj <- createGiottoObject(GetAssayData(d2, layer="counts"),
                          norm_expr = GetAssayData(d2),
                          spatial_locs = d2@reductions$spatial@cell.embeddings,
                          cell_metadata = d2@meta.data, instructions = instr)
```

## Create spatial networks (Delaunay) and calculate proximity enrichments

```{r}
# A1-1
a1_giotto_obj <- createSpatialNetwork(gobject = a1_giotto_obj,
                                      maximum_distance_delaunay = 200)
a1_cell_proximities_deluanay <- cellProximityEnrichment(gobject = a1_giotto_obj,
                                     cluster_column = 'annotationSave',
                                     spatial_network_name = 'Delaunay_network',
                                     adjust_method = 'fdr',
                                     number_of_simulations = 1000,
                                     set_seed = TRUE,
                                     seed_number = 0)

# A1-2
a2_giotto_obj <- createSpatialNetwork(gobject = a2_giotto_obj,
                                      maximum_distance_delaunay = 200)
a2_cell_proximities_deluanay <- cellProximityEnrichment(gobject = a2_giotto_obj,
                                      cluster_column = 'annotationSave',
                                      spatial_network_name = 'Delaunay_network',
                                      adjust_method = 'fdr',
                                      number_of_simulations = 1000,
                                      set_seed = TRUE,
                                      seed_number = 0)

# C2
c2_giotto_obj <- createSpatialNetwork(gobject = c2_giotto_obj,
                                      maximum_distance_delaunay = 200)
c2_cell_proximities_deluanay <- cellProximityEnrichment(gobject = c2_giotto_obj,
                                      cluster_column = 'annotationSave',
                                      spatial_network_name = 'Delaunay_network',
                                      adjust_method = 'fdr',
                                      number_of_simulations = 1000,
                                      set_seed = TRUE,
                                      seed_number = 0)

# D2-1
d2_giotto_obj <- createSpatialNetwork(gobject = d2_giotto_obj, 
                                      maximum_distance_delaunay = 200)
d2_cell_proximities_deluanay <- cellProximityEnrichment(gobject = d2_giotto_obj,
                                      cluster_column = 'annotationSave',
                                      spatial_network_name = 'Delaunay_network',
                                      adjust_method = 'fdr',
                                      number_of_simulations = 1000,
                                      set_seed = TRUE,
                                      seed_number = 0)
```

### Calculate the average proximity enrichment scores

```{r}
avg_delaunay = list(a1_cell_proximities_deluanay$enrichm_res, 
                      a2_cell_proximities_deluanay$enrichm_res,
                      c2_cell_proximities_deluanay$enrichm_res, 
                      d2_cell_proximities_deluanay$enrichm_res) %>%
    reduce(full_join, by='unified_int')
avg_delaunay$enrichm = avg_delaunay %>% select(matches('enrichm')) %>%
  rowMeans(na.rm=T)
```

## Cell proximity visualization

```{r}
source('plot_cellproximityenrichment.R')
```

#### Average proximity

The proximity enrichment is visualized. The enrichment scores as well as the number of interactions for the averaged enrichment are reported here. The below shows Figure 3.1 in the dissertation.

```{r}
cellProximityBarplot_KC(avg_delaunay, average=T)
```

The below are visualizations for each of the four datasets.

#### A1-1
```{r}
cellProximityBarplot_KC(a1_cell_proximities_deluanay$enrichm_res)
```

#### A1-2
```{r}
cellProximityBarplot_KC(a2_cell_proximities_deluanay$enrichm_res)
```

#### C2
```{r}
cellProximityBarplot_KC(c2_cell_proximities_deluanay$enrichm_res)
```

#### D2-1
```{r}
cellProximityBarplot_KC(d2_cell_proximities_deluanay$enrichm_res)
```

## Check variability between datasets

Process the enrichment result so that only the desired values remain.
Here, the p-values of each enrichment can also be seen.

```{r}
# A1-1
## Filter cells that interact with KCs
a1_en1 = a1_cell_proximities_deluanay$enrichm_res %>% 
  filter(grepl('KCs--', unified_int)) %>% 
  separate(unified_int, c('receiver', 'sender'), sep='--')
a1_en2 = a1_cell_proximities_deluanay$enrichm_res %>% 
  filter(grepl('--KCs', unified_int)) %>% 
  separate(unified_int, c('sender', 'receiver'), sep='--')
## Only keep the sender cell type and enrichment score
a1_en = rbind(a1_en1, a1_en2[a1_en2$sender!='KCs',]) 
# View full enrichment matix
a1_en

## Filter for significant p-values
a1_p = a1_en[a1_en$p_higher_orig <= 0.05 | a1_en$p_lower_orig <= 0.05,]
# View the significant enrichment matrix
a1_p

## Select for sender and scores
a1_en = a1_en %>% select(sender, enrichm)
```

```{r}
# A1-2
## Filter cells that interact with KCs
a2_en1 = a2_cell_proximities_deluanay$enrichm_res %>% 
  filter(grepl('KCs--', unified_int)) %>% 
  separate(unified_int, c('receiver', 'sender'), sep='--')
a2_en2 = a2_cell_proximities_deluanay$enrichm_res %>% 
  filter(grepl('--KCs', unified_int)) %>% 
  separate(unified_int, c('sender', 'receiver'), sep='--')
## Only keep the sender cell type and enrichment score
a2_en = rbind(a2_en1, a2_en2[a2_en2$sender!='KCs',]) 
# View full enrichment matix
a2_en

## Filter for significant p-values
a2_p = a2_en[a2_en$p_higher_orig <= 0.05 | a2_en$p_lower_orig <= 0.05,]
# View the significant enrichment matrix
a2_p

## Select for sender and scores
a2_en = a2_en %>% select(sender, enrichm)
```

```{r}
# C2
## Filter cells that interact with KCs
c2_en1 = c2_cell_proximities_deluanay$enrichm_res %>% 
  filter(grepl('KCs--', unified_int)) %>% 
  separate(unified_int, c('receiver', 'sender'), sep='--')
c2_en2 = c2_cell_proximities_deluanay$enrichm_res %>% 
  filter(grepl('--KCs', unified_int)) %>% 
  separate(unified_int, c('sender', 'receiver'), sep='--')
## Only keep the sender cell type and enrichment score
c2_en = rbind(c2_en1, c2_en2[c2_en2$sender!='KCs',])
# View full enrichment matix
c2_en

## Filter for significant p-values
c2_p = c2_en[c2_en$p_higher_orig <= 0.05 | c2_en$p_lower_orig <= 0.05,]
# View the significant enrichment matrix
c2_p

## Select for sender and scores
c2_en = c2_en  %>% select(sender, enrichm)
```

```{r}
# D2-1
## Filter cells that interact with KCs
d2_en1 = d2_cell_proximities_deluanay$enrichm_res %>% 
  filter(grepl('KCs--', unified_int)) %>% 
  separate(unified_int, c('receiver', 'sender'), sep='--')
d2_en2 = d2_cell_proximities_deluanay$enrichm_res %>% 
  filter(grepl('--KCs', unified_int)) %>% 
  separate(unified_int, c('sender', 'receiver'), sep='--')
## Only keep the sender cell type and enrichment score
d2_en = rbind(d2_en1, d2_en2[d2_en2$sender!='KCs',]) 
# View full enrichment matix
d2_en

## Filter for significant p-values
d2_p = d2_en[d2_en$p_higher_orig <= 0.05 | d2_en$p_lower_orig <= 0.05,]
# View the significant enrichment matrix
d2_p

## Select for sender and scores
d2_en=d2_en%>% select(sender, enrichm)
```

Prepare to plot the enrichment scores.

```{r}
# Combine the four dataframes into one
delaunay_box = list(a1_en, a2_en, c2_en, d2_en) %>% 
  reduce(full_join, by='sender')
# Set row and column names
rownames(delaunay_box) = delaunay_box$sender
delaunay_box$sender = NULL
colnames(delaunay_box) = c('A1','A2','C2','D2')
# Format the dataframe for plot
delaunay_box_plot = delaunay_box %>% melt() %>%
  dplyr::mutate(Dataset=variable, Score=value)
```

The below is the table used for Table 3.1 in the dissertation.
```{r}
thesis_tbl = delaunay_box
thesis_tbl$Average = rowMeans(thesis_tbl, na.rm = T)
thesis_tbl$SD = apply(thesis_tbl, 1, sd, na.rm=TRUE)
thesis_tbl %>% arrange(-Average)
```

Create a violin/box plot of enrichment score for each dataset.
This is the Figure B.1 in the dissertation.
```{r}
fig1 = ggplot(delaunay_box_plot, aes(x=variable, y=value))+
  geom_violin(width=1)+
  geom_boxplot(width=0.2, color='black')+
  scale_fill_viridis(discrete=TRUE)+
  labs(y='Enrichment score')+
  theme(axis.title.x=element_blank())+
  scale_x_discrete(labels=c('A1-1','A1-2','C2','D2-1'))
fig1
```
