---
title: "Spatial NicheNet"
output: html_document
date: "2024-07-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

This is the main file for spatially informed NicheNet analysis.

# Load libraries

```{r}
library(nichenetr)
library(Seurat)
library(SeuratObject)
library(tidyverse)
library(dplyr)
```

# Background
## Read data
```{r}
source('data_cleanup.R')
sc = sc_clean()
sc = SetIdent(sc, value = "annot_fine_zonated")
```

```{r}
spatial_data = spatial_clean()
a1 = spatial_data[[1]]
Idents(a1) <- a1$annotationSave
a2 = spatial_data[[2]]
Idents(a2) <- a2$annotationSave
c2 = spatial_data[[3]]
Idents(c2) <- c2$annotationSave
d2 = spatial_data[[4]]
Idents(d2) <- d2$annotationSave
```

```{r}
# Get the common cell types
sender = intersect(sc$annot_fine_zonated, a1$annotationSave)
sc = subset(sc, idents=sender)
```

## Set networks

The networks and matrix for mice are used.

```{r}
lr_network = readRDS(
  url("https://zenodo.org/record/7074291/files/lr_network_mouse_21122021.rds"))
ligand_target_matrix = readRDS(url(
  "https://zenodo.org/record/7074291/files/ligand_target_matrix_nsga2r_final_mouse.rds"))
weighted_networks = readRDS(url(
  "https://zenodo.org/record/7074291/files/weighted_networks_nsga2r_final_mouse.rds"))
```

```{r}
lr_network = lr_network %>% distinct(from, to)
weighted_networks_lr = weighted_networks$lr_sig %>% 
  inner_join(lr_network, by = c("from","to"))
```

# Gene set of interest

```{r}
# Set receiver as kupffer cells
receiver = "KCs"
expressed_genes_receiver = get_expressed_genes(receiver, sc, pct = 0.10)
```

```{r}
# Set background gene set as the whole genome
background_expressed_genes = sc@assays[["RNA"]]@counts@Dimnames[[1]] %>% 
  unique() %>% .[. %in% rownames(ligand_target_matrix)]
```

```{r}
# Sender genes
sender_celltypes = sender[sender != receiver]

list_expressed_genes_sender = sender_celltypes %>% unique() %>% 
  lapply(get_expressed_genes, sc, 0.10)
expressed_genes_sender = list_expressed_genes_sender %>% unlist() %>% unique()
```

## Find gene set of interest
```{r}
# Find the gene set of interest
seurat_obj_receiver = subset(sc, idents = receiver)
seurat_obj_receiver = SetIdent(seurat_obj_receiver, value =
                        seurat_obj_receiver[["annot_fine_zonated", drop=TRUE]])
```

```{r}
# KCs against MoMac1 and 2
DE_table_receiver = FindMarkers(object = sc, ident.1 = receiver, 
                          ident.2 = c('MoMac1', 'MoMac2'), min.pct = 0.10) %>%
  rownames_to_column("gene")
```

```{r}
# Find gene set of interest
geneset_oi = DE_table_receiver %>% 
  filter(p_val_adj <= 0.05 & avg_log2FC >= 0.25) %>% pull(gene)
geneset_oi = geneset_oi %>% .[. %in% rownames(ligand_target_matrix)]
```

# Ligand activities

```{r}
ligands = lr_network %>% pull(from) %>% unique()
receptors = lr_network %>% pull(to) %>% unique()

expressed_ligands = intersect(ligands,expressed_genes_sender)
expressed_receptors = intersect(receptors,expressed_genes_receiver)
```

```{r}
# Define potential ligands
potential_ligands = lr_network %>% 
  filter(from %in% expressed_ligands & to %in% expressed_receptors) %>% 
  pull(from) %>% unique()
```

```{r}
# Get ligand activities
ligand_activities = predict_ligand_activities(
  geneset = geneset_oi, 
  background_expressed_genes = background_expressed_genes,
  ligand_target_matrix = ligand_target_matrix, 
  potential_ligands = potential_ligands
  )
ligand_activities = ligand_activities %>% arrange(-aupr_corrected) %>% 
  mutate(rank = rank(desc(aupr_corrected)))
```

```{r}
# Reformat the network for further analysis
lr_network_renamed = lr_network %>% rename(ligand=from, receptor=to)
```

# Obtain the DE and expression tables

```{r}
# Make DE and expression info tables
DE_table = calculate_de(sc, celltype_colname = "annot_fine_zonated",
                     features = union(expressed_ligands, expressed_receptors))
expression_info = get_exprs_avg(sc, "annot_fine_zonated")


processed_DE_table = process_table_to_ic(DE_table, 
                                         table_type = "celltype_DE",
                                         lr_network_renamed, 
                                         senders_oi = sender_celltypes, 
                                         receivers_oi = receiver)

processed_expr_table = process_table_to_ic(expression_info, 
                                            table_type = "expression", 
                                            lr_network_renamed)
```

```{r}
# This line is necessary as the process_table_to_ic function
# mutates the underscores in cell type annotations to dashes
processed_expr_table = processed_expr_table %>% 
  mutate(across(c('sender','receiver'), \(x) str_replace(x, '-','_')))
```

# Get prioritization tables

## Run original NicheNet
```{r}
# Weights for original NicheNet
prioritizing_weights = c("de_ligand" = 1,
                         "de_receptor" = 1,
                         "activity_scaled" = 2,
                         "exprs_ligand" = 1,
                         "exprs_receptor" = 1,
                         "ligand_condition_specificity" = 0,
                         "receptor_condition_specificity" = 0)

# Generate prioritization table (NicheNet)
prior_table = generate_prioritization_tables(
  processed_expr_table,
  processed_DE_table,
  ligand_activities,
  prioritizing_weights = prioritizing_weights
  )
prior_table
```

## Run spatial NicheNet
```{r}
source('prioritization.R')
source('plot_prior.R')
# Weights for spatial NicheNet
prioritizing_weights_spatial = c("de_ligand" = 1,
                                 "de_receptor" = 1,
                                 "activity_scaled" = 2,
                                 "exprs_ligand" = 1,
                                 "exprs_receptor" = 1,
                                 "ligand_condition_specificity" = 0,
                                 "receptor_condition_specificity" = 0,
                                 "spatial_proximity" = 1)
```

### Giotto

```{r}
source('giotto_fin.R')
proximity_matrices = giotto_spatial_proximity(a1,a2,c2,d2)
```

#### Prioritization using each spatial data

```{r}
a1_proximity = proximity_matrices[2]$enrichm_res
a2_proximity = proximity_matrices[4]$enrichm_res
c2_proximity = proximity_matrices[6]$enrichm_res
d2_proximity = proximity_matrices[8]$enrichm_res
```

##### A1-1 
```{r}
spatial_prior_a1 = scaled_prioritization_table(
  processed_expr_table,
  processed_DE_table,
  ligand_activities,
  spatial_matrix = a1_proximity,
  prioritizing_weights = prioritizing_weights_spatial)

# Prioritizaion table
spatial_prior_a1

# Prioritzation changes compared to original NicheNet
a1_vs_orginal = plot_prioritization(prior_table, spatial_prior_a1)
a1_vs_orginal
```

##### A1-2
```{r}
spatial_prior_a2 = scaled_prioritization_table(
  processed_expr_table,
  processed_DE_table,
  ligand_activities,
  spatial_matrix = a2_proximity,
  prioritizing_weights = prioritizing_weights_spatial)

# Prioritizaion table
spatial_prior_a2

# Prioritzation changes compared to original NicheNet
a2_vs_orginal = plot_prioritization(prior_table, spatial_prior_a2)
a2_vs_orginal
```

##### C2
```{r}
spatial_prior_c2 = scaled_prioritization_table(
  processed_expr_table,
  processed_DE_table,
  ligand_activities,
  spatial_matrix = c2_proximity,
  prioritizing_weights = prioritizing_weights_spatial)

# Prioritizaion table
spatial_prior_c2

# Prioritzation changes compared to original NicheNet
c2_vs_orginal = plot_prioritization(prior_table, spatial_prior_c2)
c2_vs_orginal
```

##### D2-1
```{r}
spatial_prior_d2 = scaled_prioritization_table(
  processed_expr_table,
  processed_DE_table,
  ligand_activities,
  spatial_matrix = d2_proximity,
  prioritizing_weights = prioritizing_weights_spatial)

# Prioritizaion table
spatial_prior_d2

# Prioritzation changes compared to original NicheNet
d2_vs_orginal = plot_prioritization(prior_table, spatial_prior_d2, add_col=T)
d2_vs_orginal
```

#### Prioritization using average enrichment

```{r}
# Average enrichment score from the four datasets
avg_delaunay = as.data.frame(proximity_matrices[9:50])
```

```{r}
spatial_prior_full = scaled_prioritization_table(
  processed_expr_table,
  processed_DE_table,
  ligand_activities,
  spatial_matrix = avg_delaunay,
  prioritizing_weights = prioritizing_weights_spatial)

# Prioritizaion table
spatial_prior_full

# Prioritzation changes compared to original NicheNet
giotto_vs_orginal = plot_prioritization(prior_table, spatial_prior_full)
giotto_vs_orginal
```

```{r}
# Plot the variability of the results
source('variability_plot.R')
figures = plot_variability_ranking(spatial_prior_full, 
                                   spatial_prior_a1,
                                   spatial_prior_a2,
                                   spatial_prior_c2,
                                   spatial_prior_d2)
figures[[2]]
```

##### Prioritization without scaling

```{r}
spatial_prior_unscaled = unscaled_prioritization_table(
  processed_expr_table,
  processed_DE_table,
  ligand_activities,
  spatial_matrix = avg_delaunay,
  prioritizing_weights = prioritizing_weights_spatial)

# Prioritizaion table
spatial_prior_unscaled

# Prioritzation changes compared to original NicheNet
giottoU_vs_orginal = plot_prioritization(prior_table, spatial_prior_unscaled)
giottoU_vs_orginal
```

```{r}
# Plot the variability of the results
figures = plot_variability_ranking(spatial_prior_unscaled, 
                                   spatial_prior_a1,
                                   spatial_prior_a2,
                                   spatial_prior_c2,
                                   spatial_prior_d2)
figures[[2]]
```

##### Prioritization with different weight
```{r}
spatial_prior_weight = unscaled_prioritization_table(
  processed_expr_table,
  processed_DE_table,
  ligand_activities,
  spatial_matrix = avg_delaunay,
  prioritizing_weights = c("de_ligand" = 1,
                                 "de_receptor" = 1,
                                 "activity_scaled" = 2,
                                 "exprs_ligand" = 1,
                                 "exprs_receptor" = 1,
                                 "ligand_condition_specificity" = 0,
                                 "receptor_condition_specificity" = 0,
                                 "spatial_proximity" = 2.4))

# Prioritizaion table
spatial_prior_weight

# Prioritzation changes compared to original NicheNet
giottoW_vs_orginal = plot_prioritization(prior_table, spatial_prior_weight)
giottoW_vs_orginal
```

```{r}
# Plot the variability of the results
figures = plot_variability_ranking(spatial_prior_weight, 
                                   spatial_prior_a1,
                                   spatial_prior_a2,
                                   spatial_prior_c2,
                                   spatial_prior_d2)
figures[[2]]
```


#### Differential distance
```{r}
source('giotto_fin.R')
source('prioritization.R')
communication_giotto = giotto_diff(a1)
juxtacrine = communication_giotto[[1]]$enrichm_res
paracrine = communication_giotto[[2]]$enrichm_res
endocrine = communication_giotto[[3]]$enrichm_res
```

```{r}
spatial_prior_dist = lr_diff_prioritization_table(
  processed_expr_table,
  processed_DE_table,
  ligand_activities,
  spatial_matrix_connected = juxtacrine,
  spatial_matrix_paracrine = endocrine,
  prioritizing_weights = prioritizing_weights_spatial)

# Prioritizaion table
spatial_prior_dist

# Distance-based Giotto vs original
dist_vs_original = plot_prioritization(prior_table, spatial_prior_dist)
dist_vs_original
```

### Prioritization using Squidpy enrichment

Read in the enrichment scores and preprocess them.

```{r}
a1_sq = read.csv('data/A1_1_sq.csv') %>%
  mutate(enrichm = enrichment.depletion, sender=Sender) %>%
  select('sender', 'count', 'enrichm')
a2_sq = read.csv('data/A1_2_sq.csv') %>%
  mutate(enrichm = enrichment.depletion, sender=Sender) %>%
  select('sender', 'count', 'enrichm')
c2_sq = read.csv('data/C2_sq.csv') %>%
  mutate(enrichm = enrichment.depletion, sender=Sender) %>%
  select('sender', 'count', 'enrichm')
d2_sq = read.csv('data/D2_1_sq.csv') %>%
  mutate(enrichm = enrichment.depletion, sender=Sender) %>%
  select('sender', 'count', 'enrichm')
```

#### Prioritization using each spatial data

##### A1-1
```{r}
spatial_prior_sq_a1 = squidpy_prioritization_table(
  processed_expr_table,
  processed_DE_table,
  ligand_activities,
  spatial_matrix = a1_sq,
  prioritizing_weights = prioritizing_weights_spatial)
spatial_prior_sq_a1

sqA1_vs_original = plot_prioritization(prior_table, spatial_prior_sq_a1)
sqA1_vs_original
```
##### A1-2
```{r}
spatial_prior_sq_a2 = squidpy_prioritization_table(
  processed_expr_table,
  processed_DE_table,
  ligand_activities,
  spatial_matrix = a2_sq,
  prioritizing_weights = prioritizing_weights_spatial)
spatial_prior_sq_a2

sqa2_vs_original = plot_prioritization(prior_table, spatial_prior_sq_a2)
sqa2_vs_original
```

##### C2
```{r}
spatial_prior_sq_c2 = squidpy_prioritization_table(
  processed_expr_table,
  processed_DE_table,
  ligand_activities,
  spatial_matrix = c2_sq,
  prioritizing_weights = prioritizing_weights_spatial)
spatial_prior_sq_c2

sqc2_vs_original = plot_prioritization(prior_table, spatial_prior_sq_c2)
sqc2_vs_original
```

##### D2-1
```{r}
spatial_prior_sq_d2 = squidpy_prioritization_table(
  processed_expr_table,
  processed_DE_table,
  ligand_activities,
  spatial_matrix = d2_sq,
  prioritizing_weights = prioritizing_weights_spatial)
spatial_prior_sq_d2

sqd2_vs_original = plot_prioritization(prior_table, spatial_prior_sq_d2, 
                                       add_col=T)
sqd2_vs_original
```
#### Plot variability of the results

```{r}
figures = plot_variability_ranking_sq(spatial_prior_sq_a1,
                                      spatial_prior_sq_a2,
                                      spatial_prior_sq_c2,
                                      spatial_prior_sq_d2)

figures[[2]]
```
